A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods

### 2025-04-14 2025/RuSCDays/RSD-2025.tex
some fix

\keywords{Global optimization  \and Multiextremal Functions \and Partially Defined Functions \and Hyperparameter Tuning \and  Time Series \and Artificial Intelligence.}
\end{abstract}
%
%
%

@@

\begin{equation}\label{rule_points_index} 
v_i=v(x_i)=
\begin{cases}
    0     & \quad \text{if } x_i \text{ is boundary point},\\
    1     & \quad \text{if } x_i \text{ is internal point}.
\end{cases}
\end{equation}
%(рус.) Точки x_0=0 и x_(k+p)=1 введены дополнительно (значения z_0 и z_(k+p) не определены) для удобства последующих обозначений.
The points $x_0=0$ and $x_{k+p}=1$ are additionally introduced (the values $z_0$ and $z_{k+p}$ are not defined) for the convenience of subsequent designations.

@@

%(рус.) 4. Каждому интервалу (x_(i-1),x_i ),x_(i-1)  ∉ S_k,x_i  ∉ S_k,2≤ i≤ k+p, поставить в соответствие число R(i), которое называется характеристикой интервала и вычисляется по формуле
\item Assign a number $R(i)$ to each interval $(x_{i-1}, x_i), x_(i-1) \notin S_k, x_i \notin S_k, 2 \leq i \leq k+p$, which is called the interval characteristic and is calculated using the formula
\begin{equation}\label{characteristic}
R(i)=
\begin{cases}
    \Delta_i+\frac{(z_i-z_{i-1})^2}{(r\mu)^2\Delta_i} - 2 \frac{(z_i+z_{i-1}-2z^*)}{r\mu},   & \quad v(x_i)=v(x_{i-1})=1,\\
    2\Delta_i- 4 \frac{(z_i-z^*)}{r\mu},   & \quad v(x_i)=1, v(x_{i-1})=0,\\
    2\Delta_i- 4 \frac{(z_{i-1}-z^*)}{r\mu},       & \quad v(x_i)=0, v(x_{i-1})=1,\\
\end{cases}
\end{equation}
%(рус.) где Δ_i  =(x_i-x_(i-1) )^(1/N) , and r>1 – параметр алгоритма.
where $\Delta_i=(x_i-x_{i-1})^\frac{1}{N}$, and $r>1$ is the parameter of the algorithm.

@@

%(рус.) Отметим, что функция f(x), вообще говоря, может иметь разрывы в целочисленных точках x_i  = i. Так как эти точки известны заранее, их можно на первой итерации добавить в информационную базу алгоритма со значением индекса равным 0. В соответствии с правилами алгоритма, в процессе его работы не будет возникать интервалов, обе граничные точки которых имеют нулевой индекс, и правила вычисления характеристик (7) и точки нового интервала (8)-(9) будут валидными.
Note that the function $f(x)$, in general, can have discontinuities at integer points $x_i=i$. Since these points are known in advance, they can be added to the information base of the algorithm at the first iteration with an index value equal to $0$. According to the rules of the algorithm, during its operation there will be no intervals with both boundary points having zero index, and the rules for calculating the features (\ref{characteristic}) and the points of the new interval (\ref{new_point_1})-(\ref{new_point_2}) will be valid.

@@

%(рус.) Также требуется дополнить правило выбора точки нового испытания для обработки ситуаций, когда обе граничные точки имеют нулевой индекс:
It is also necessary to supplement the rule for selecting a new trial point to handle situations where both boundary points have a zero index:
\begin{equation}\label{new_point_1_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t) \text{ or } v(x_{t-1})=v(x_t)=0,
\end{equation}
\begin{equation}\label{new_point_2_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{\mu} \right]^N,
\end{equation}
if $v(x_{t-1})=v(x_t) = 1$.

@@

%(рус.) У lagged-преобразования имеется один параметр window_size, который изменяется от 5 до 250 с шагом 1. Поскольку этот параметр имеет большое количество значений, при оптимизации, будем рассматривать его как непрерывный (с округлением до единиц).
The lagged transformation has one parameter $window\_size$, which varies from 5 to 250 in steps of 1. Since this parameter has a large number of values, we will consider it as continuous (rounded to units) during optimization.

>>>>

\keywords{Global Optimization  \and Multiextremal Functions \and Partially Defined Functions \and Hyperparameter Tuning \and  Time Series \and Artificial Intelligence.}
\end{abstract}
%
%
%

@@

\begin{equation}\label{rule_points_index} 
v_i=v(x_i)=
\begin{cases}
    0,     & \quad \text{if } x_i \text{ is boundary point},\\
    1,     & \quad \text{if } x_i \text{ is internal point}.
\end{cases}
\end{equation}
%(рус.) Точки x_0=0 и x_(k+p)=1 введены дополнительно (значения z_0 и z_(k+p) не определены) для удобства последующих обозначений.
The points $x_0=0$ and $x_{k+p}=1$ are additionally introduced (the values $z_0$ and $z_{k+p}$ are not defined) for the convenience of subsequent designations.

@@

%(рус.) 4. Каждому интервалу (x_(i-1),x_i ),x_(i-1)  ∉ S_k,x_i  ∉ S_k,2≤ i≤ k+p, поставить в соответствие число R(i), которое называется характеристикой интервала и вычисляется по формуле
\item Assign a number $R(i)$ to each interval $(x_{i-1}, x_i), x_{i-1} \notin S_k, x_i \notin S_k, 2 \leq i \leq k+p$, which is called the interval characteristic and is calculated using the formula
\begin{equation}\label{characteristic}
R(i)=
\begin{cases}
    \Delta_i+\frac{(z_i-z_{i-1})^2}{(r\mu)^2\Delta_i} - 2 \frac{(z_i+z_{i-1}-2z^*)}{r\mu},   & \quad v(x_i)=v(x_{i-1})=1,\\
    2\Delta_i- 4 \frac{(z_i-z^*)}{r\mu},   & \quad v(x_i)=1, v(x_{i-1})=0,\\
    2\Delta_i- 4 \frac{(z_{i-1}-z^*)}{r\mu},       & \quad v(x_i)=0, v(x_{i-1})=1,\\
\end{cases}
\end{equation}
%(рус.) где Δ_i  =(x_i-x_(i-1) )^(1/N) , and r>1 – параметр алгоритма.
where $\Delta_i=(x_i-x_{i-1})^\frac{1}{N}$, and $r>1$ is the parameter of the algorithm.

@@

%(рус.) Отметим, что функция f(x), вообще говоря, может иметь разрывы в целочисленных точках x_i  = i. Так как эти точки известны заранее, их можно на первой итерации добавить в информационную базу алгоритма со значением индекса равным 0. В соответствии с правилами алгоритма, в процессе его работы не будет возникать интервалов, обе граничные точки которых имеют нулевой индекс, и правила вычисления характеристик (7) и точки нового интервала (8)-(9) будут валидными.
Note that the function $f(x)$, in general, can have discontinuities at integer points $x_i=i$. Since these points are known in advance, they can be added to the information base of the algorithm at the first iteration with an index value equal to $0$. According to the rules of the algorithm, during its operation there will be no intervals with both boundary points having zero index, and the rules for calculating the characteristics (\ref{characteristic}) and the points of the new interval (\ref{new_point_1})-(\ref{new_point_2}) will be valid.

@@

%(рус.) Также требуется дополнить правило выбора точки нового испытания для обработки ситуаций, когда обе граничные точки имеют нулевой индекс:
It is also necessary to supplement the rule for selecting a new trial point to handle situations where both boundary points have a zero index:
\begin{equation}\label{new_point_1_2} 
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}, \text{ if } v(x_{t-1}) \neq v(x_t) \text{ or } v(x_{t-1})=v(x_t)=0,
\end{equation}
\begin{equation}\label{new_point_2_2} 
\begin{split}
x^{k+p+1} = \frac{x_t+x_{t-1}}{2}-\frac{\textit{sign}{(z_t-z_{t-1})}}{2r} \left[ \frac{|z_t-z_{t-1}|}{\mu} \right]^N, \\
\text{if }  v(x_{t-1})=v(x_t) = 1.
\end{split}
\end{equation}

@@

%(рус.) У lagged-преобразования имеется один параметр window_size, который изменяется от 5 до 250 с шагом 1. Поскольку этот параметр имеет большое количество значений, при оптимизации, будем рассматривать его как непрерывный (с округлением до единиц).
The lagged transformation has one parameter window\_size, which varies from 5 to 250 in steps of 1. Since this parameter has a large number of values, we will consider it as continuous (rounded to units) during optimization.
